---
title: "GPPD Diagnostic Report"
author: "Sofia Perondi"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    toc: true
    toc_float: true
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
```

This report provides a diagnostic analysis of the Global Public Procurement Dataset (GPPD) filtered for EU-27 countries, covering contracts awarded between 2006 and 2021.
The dataset is contract-level and includes information on the contracting authority (buyer), suppliers (bidders), contract values, sector of activity, and selected indicators of process quality (e.g. single bidding, selection method).

``` 
# Load dataset (already cleaned and saved)
gppd_complete <- fread("gppd_outsourcing_data_FINAL.csv")

cat("Final dataset contains", nrow(gppd_complete), "contracts\n")
cat("Countries included:", toString(sort(unique(gppd_complete$buyer_country))), "\n")
```

We first evaluate the extent of missing values across variables. This is crucial to decide which variables can be used robustly in empirical models.

```{r}
missingness <- sapply(gppd_complete, function(x) mean(is.na(x)) * 100)
missingness <- sort(missingness, decreasing = TRUE)
head(missingness, 10)
```

Interpretation:
Variables like lot_updateddurationdays and bid_subcontractedproportion are missing almost entirely (>90%), and therefore should not be used in the core analysis.
Procurement process indicators such as corr_singleb (single bidding) have moderate missingness (~20%) but remain usable.
Core variables (buyer_country, buyer_nuts, tender_year, tender_supplytype, tender_digiwhist_price) are fully populated and reliable.

```{r}
table(gppd_complete$tender_supplytype, useNA = "ifany")
```

Contracts are classified into three main categories (Works, Supplies, Services) according to the harmonised supply type.

```{r}
summary_supplytype <- gppd_complete %>%
  group_by(tender_supplytype) %>%
  summarise(
    n_contracts = n(),
    total_value_billion = sum(tender_digiwhist_price, na.rm = TRUE) / 1e9,
    avg_value_million = mean(tender_digiwhist_price, na.rm = TRUE) / 1e6,
    .groups = "drop"
  )
summary_supplytype
```

Supplies dominate in terms of number of contracts and total value.
Services show high diversity but lower average values.
Works contracts are fewer but often represent large-scale infrastructure projects.

```{r}
summary_country_year <- gppd_complete %>%
  group_by(buyer_country, tender_year) %>%
  summarise(
    n_contracts = n(),
    total_value_billion = sum(tender_digiwhist_price, na.rm = TRUE) / 1e9,
    .groups = "drop"
  )
head(summary_country_year, 27)
```
This table shows the evolution of contract counts and values at the country-year level, which will be the main unit for later econometric analysis.

Not all countries have full coverage for the entire 2006–2021 period. We inspect which years are present for each country.

```{r}
coverage <- gppd_complete %>%
  group_by(buyer_country, tender_year) %>%
  summarise(n_contracts = n(), .groups = "drop") %>%
  complete(buyer_country, tender_year = 2006:2021, fill = list(n_contracts = 0))

coverage_summary <- coverage %>%
  group_by(buyer_country) %>%
  summarise(
    first_year = min(tender_year[n_contracts > 0], na.rm = TRUE),
    last_year  = max(tender_year[n_contracts > 0], na.rm = TRUE),
    total_years = sum(n_contracts > 0),
    .groups = "drop"
  )
coverage_summary
```

For many countries, reliable data begin only around 2011–2014.
Some countries (e.g. Romania) have longer coverage but not necessarily with consistent reporting quality.

```{r}
coverage_matrix <- coverage %>%
  mutate(has_data = ifelse(n_contracts > 0, 1, 0))

ggplot(coverage_matrix, aes(x = tender_year, y = buyer_country, fill = factor(has_data))) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("0" = "lightgrey", "1" = "steelblue"),
                    name = "Data available",
                    labels = c("No", "Yes")) +
  labs(title = "Coverage of GPPD Contracts by Country and Year",
       x = "Year", y = "Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We compare two possible panels:
Full panel (2006–2021): maximises years, but many early years are empty.
Comparable panel (2011–2021): shorter period, but more consistent across EU-27.

```{r}
panel_unbalanced <- gppd_complete %>%
  group_by(buyer_country, tender_year) %>%
  summarise(
    n_contracts = n(),
    total_value_billion = sum(tender_digiwhist_price, na.rm = TRUE) / 1e9,
    .groups = "drop"
  ) %>%
  complete(buyer_country, tender_year = 2006:2021,
           fill = list(n_contracts = 0, total_value_billion = 0))

panel_balanced <- panel_unbalanced %>%
  filter(tender_year >= 2011 & tender_year <= 2021)

compare_loss <- data.frame(
  Dataset = c("Full 2006–2021", "Comparable 2011–2021"),
  Years = c(length(unique(panel_unbalanced$tender_year)),
            length(unique(panel_balanced$tender_year))),
  Total_contracts = c(sum(panel_unbalanced$n_contracts),
                      sum(panel_balanced$n_contracts)),
  Total_value_B = c(sum(panel_unbalanced$total_value_billion),
                    sum(panel_balanced$total_value_billion))
)
compare_loss
```

Restricting to 2011–2021 removes ~7% of contracts but only ~0.05% of total value.
This suggests early years are sparse and low-value relative to the bulk of the dataset.

Conclusion:
Coverage: EU-27 dataset is rich but uneven; reliable coverage starts mostly around 2011.
Variables: Core contract values and sector classifications are robust; process indicators have partial availability; subcontracting data are largely unusable.
Panel choice: Both full (2006–2021) and comparable (2011–2021) panels are possible. For econometric modelling, the shorter window may be preferable for consistency.

```{r}
# install.packages("xtable")
library(xtable)

# Export coverage_summary as .tex
tab_cov <- xtable(coverage_summary,
                  caption = "Coverage of GPPD Contracts by Country (2006–2021)",
                  label = "tab:gppd_coverage")

print(tab_cov, file = "coverage_summary.tex", include.rownames = FALSE)
```