---
title: "Full Model – Advanced Analysis"
author: "Sofia Perondi"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: false
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{mathtools}
  - \setlength{\emergencystretch}{2em}
---

```{r setup, include=FALSE}
library(tidyverse)
library(fixest)
library(modelsummary)
library(kableExtra)
library(broom)

knitr::opts_chunk$set(
echo = FALSE,
message = FALSE,
warning = FALSE,
results = "asis"
)

dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)
dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
base_path <- "~/Desktop/Tesina IV /Data analysis /Downloaded data"

df_raw <- read_csv(
file.path(base_path, "FINAL_ANALYSIS_DATASET_2006_2021.csv"),
show_col_types = FALSE
)

df_country <- df_raw %>%
group_by(buyer_country, tender_year) %>%
summarise(
avg_contract_price     = mean(tender_digiwhist_price, na.rm = TRUE),
total_expenditure_real = mean(total_expenditure_real, na.rm = TRUE),
unemp_rate             = mean(unemp_rate, na.rm = TRUE),
pop_density            = mean(pop_density, na.rm = TRUE),
gov_debt_pct_gdp       = mean(gov_debt_pct_gdp, na.rm = TRUE),
corruption_control     = mean(corruption_control, na.rm = TRUE),
gov_state_market       = mean(gov_state_market, na.rm = TRUE),
recession              = mean(recession, na.rm = TRUE),
n_contracts_year       = n(),
.groups = "drop"
)

df <- df_country %>%
filter(tender_year >= 2011, tender_year <= 2021) %>%
filter(total_expenditure_real > 0, avg_contract_price > 0) %>%
mutate(
ln_spesa       = log(total_expenditure_real),
ln_valmean     = log(avg_contract_price),
ln_unemp_rate  = log(unemp_rate),
ln_pop_density = log(pop_density),
ln_gov_debt    = log(gov_debt_pct_gdp)
) %>%
group_by(buyer_country) %>%
mutate(
ln_spesa_dm = ln_spesa - mean(ln_spesa, na.rm = TRUE),
X           = ln_valmean - mean(ln_valmean, na.rm = TRUE)
) %>%
ungroup()
```

# Pre vs Post-2016 Two-way Fixed Effects and Interaction Test

This section evaluates whether the relationship between outsourcing intensity and public expenditure changes after 2016, the year in which the EU Procurement Directives were transposed and transparency requirements were strengthened.

Three specifications are estimated:

1. A two-way fixed effects model on the pre-2016 subsample.
2. The same specification on the post-2016 subsample.
3. A full-sample TWFE model including an interaction term between outsourcing intensity and a post-2016 dummy.

The estimated equation for the interaction model is:

\[
\ln(\text{expenditure}_{ct})^{dm}
= \beta X_{ct}
+ \delta \left(X_{ct} \cdot \text{Post2016}_t\right)
+ \gamma Z_{ct}
+ \alpha_c + \lambda_t + \varepsilon_{ct}
\]

where:

- \(X_{ct}\) is demeaned log outsourcing intensity,
- \(Z_{ct}\) are log-transformed macro controls (plus corruption, state–market, recession),
- \(\alpha_c\) are country fixed effects,
- \(\lambda_t\) are year fixed effects.

The parameter \(\delta\) measures whether the marginal relationship between outsourcing and expenditure changes after 2016.


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
df_pre  <- df %>% filter(tender_year < 2016)
df_post <- df %>% filter(tender_year >= 2016)

m_pre <- feols(
ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density + ln_gov_debt +
corruption_control + gov_state_market + recession |
buyer_country + tender_year,
data = df_pre, cluster = ~ buyer_country
)

m_post <- feols(
ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density + ln_gov_debt +
corruption_control + gov_state_market + recession |
buyer_country + tender_year,
data = df_post, cluster = ~ buyer_country
)

df <- df %>% mutate(post2016 = if_else(tender_year >= 2016, 1, 0))

m_inter <- feols(
ln_spesa_dm ~ X + X:post2016 + ln_unemp_rate + ln_pop_density +
ln_gov_debt + corruption_control + gov_state_market + recession |
buyer_country + tender_year,
data = df, cluster = ~ buyer_country
)

# TWFE table for pre/post 2016 + interaction

tab_prepost <- etable(
  list(
    "Pre-2016 TWFE"     = m_pre,
    "Post-2016 TWFE"    = m_post,
    "Interaction model" = m_inter
  ),
  tex = TRUE,
  digits = 4,
  cluster = "buyer_country",
  fitstat = ~ n + r2,
  title = "Two-way FE: Pre/Post 2016 and Interaction Model",
  placement = "H"   # <<< forza la posizione
)

tab_prepost

```

# Repeated Cross-sectional Estimates by Year

This section estimates a sequence of year-specific cross-sectional regressions. For each year \(t\), the following model is estimated:

\[
\ln(\text{expenditure}_{ct})
= \beta_t \ln(\text{outsourcing}_{ct})
+ \gamma_t Z_{ct}
+ \varepsilon_{ct}
\]

This specification does not include fixed effects.  
Each coefficient \(\beta_t\) captures the contemporaneous association between outsourcing intensity and public expenditure **within a given cross section**, allowing for time-variation in the magnitude and sign of the relationship.


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# --- Repeated Cross-sectional OLS (per Year) ---

cs_results <- df %>%
group_by(tender_year) %>%
group_modify(~ {
m <- lm(
ln_spesa ~ ln_valmean + ln_unemp_rate + ln_pop_density +
ln_gov_debt + corruption_control + gov_state_market + recession,
data = .x
)
tidy(m) %>% mutate(n_obs = nrow(.x))
}) %>% ungroup()

cs_X <- cs_results %>%
filter(term == "ln_valmean") %>%
mutate(ci_low = estimate - 1.96*std.error,
ci_high = estimate + 1.96*std.error)

cs_table <- kable(
  cs_X %>% select(tender_year, estimate, std.error, ci_low, ci_high, p.value, n_obs),
  format = "latex",
  booktabs = TRUE,
  caption = "Repeated cross-sectional estimates for log of valmean"
) %>%
  kable_styling(latex_options = c("HOLD_position"))  # <<< Fissa la tabella

cs_table
```

# Country-Level Correlation and Relationship With Micro-Data Coverage

The first diagnostic computes, for each country \(c\), the Pearson correlation between demeaned public expenditure and demeaned outsourcing intensity:

\[
\rho_c = \operatorname{Corr}\!\left( 
\ln(\text{expenditure}_{ct})^{dm},\, X_{ct}
\right)
\]


Since both variables are demeaned within country, \(\rho_c\) captures **purely within-country co-movement**, abstracting from long-run cross-country differences.

To evaluate whether the strength or sign of this correlation relates to the quality of the underlying procurement micro data, we construct a measure of coverage:
\[
\text{coverage}_c = \sum_t N_{ct}
\]


where \(N_{ct}\) is the number of micro-level contracts observed for country \(c\) in year \(t\).

We then compute:

- the correlation between \(\rho_c\) and coverage,
- the correlation between coverage and the indicator \(1\{\rho_c > 0\}\).

These diagnostics assess whether limited micro coverage is systematically associated with weaker or negative within-country correlations.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
cor_table <- df %>%
group_by(buyer_country) %>%
summarise(
cor_YX = cor(ln_spesa_dm, X, use = "complete.obs"),
n_obs = n(), .groups = "drop"
)

coverage <- df_raw %>%
filter(tender_year >= 2011, tender_year <= 2021) %>%
count(buyer_country) %>%
rename(total_contracts = n)

cor_cov <- cor_table %>%
left_join(coverage, by = "buyer_country") %>%
mutate(positive_corr = if_else(cor_YX > 0, 1, 0))

tab_corr <- kable(
  cor_cov,
  format = "latex",
  booktabs = TRUE,
  caption = "Within-country correlation and micro-data coverage"
) %>%
  kable_styling(latex_options = c("HOLD_position"))

tab_corr

```

## Coverage Tertiles: Group-Based Diagnostics

To investigate whether the relationship between outsourcing and expenditure varies across countries with different levels of micro-data completeness, we partition countries into three equally sized groups (tertiles) based on micro-data coverage:

- Tertile 1: countries with the lowest number of observed contracts,
- Tertile 2: intermediate coverage,
- Tertile 3: highest coverage.

For each group \(g\), we compute:

- the average number of contracts per country in the group,
- the average within-country correlation \(\rho_c\),
- the share of countries with a positive correlation,
- the number of countries in the group.

This provides a coarse but informative classification, showing whether higher-coverage countries exhibit systematically different patterns in the outsourcing–expenditure linkage.
```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
cor_cov <- cor_cov %>%
mutate(coverage_group = ntile(total_contracts, 3))

coverage_summary <- cor_cov %>%
group_by(coverage_group) %>%
summarise(
mean_coverage   = mean(total_contracts),
mean_correlation= mean(cor_YX),
share_positive  = mean(positive_corr),
n_countries     = n(),
.groups = "drop"
)

tab_cov <- kable(
  coverage_summary,
  format = "latex",
  booktabs = TRUE,
  caption = "Coverage tertiles and correlation patterns"
) %>%
  kable_styling(latex_options = c("HOLD_position"))

tab_cov
```

# TWFE Models by Expenditure Tertiles

To assess heterogeneity in the outsourcing–expenditure relationship across different levels of public spending, countries are partitioned into three groups based on the distribution of their average log real expenditure:

- Group 1: lowest average expenditure (bottom tertile),
- Group 2: medium expenditure (middle tertile),
- Group 3: highest average expenditure (top tertile).

Formally, for each country \(c\) we compute:

\[
\overline{\ln(\text{expenditure}_{c})}
= \frac{1}{T_c}
\sum_t \ln(\text{expenditure}_{ct})
\]


Countries are then assigned to tertiles using the empirical quantiles of this distribution.

For each expenditure group \(g\), we estimate a two-way fixed effects model:

\[
\ln(\text{expenditure}_{ct})^{dm}
= \beta_g X_{ct}
+ \gamma_g Z_{ct}
+ \alpha_c + \lambda_t + \varepsilon_{ct}
\]

This approach allows the coefficient of interest \(\beta_g\) to vary across low-, medium-, and high-expenditure countries, providing evidence on whether outsourcing intensity has differential associations depending on the scale of the public sector.
.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
country_mean <- df %>%
group_by(buyer_country) %>%
summarise(mean_ln_spesa = mean(ln_spesa), .groups = "drop") %>%
mutate(Y_group = ntile(mean_ln_spesa, 3))

df_q <- df %>% left_join(country_mean, by = "buyer_country")

groups <- sort(unique(df_q$Y_group))

models_q <- map(
groups,
~ feols(
ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density + ln_gov_debt +
corruption_control + gov_state_market + recession |
buyer_country + tender_year,
data = df_q %>% filter(Y_group == .x),
cluster = ~ buyer_country
)
)

names(models_q) <- paste("Group", groups)

tab_q <- etable(
  models_q,
  tex = TRUE,
  digits = 4,
  cluster = "buyer_country",
  fitstat = ~ n + r2,
  title = "TWFE by expenditure tertile",
  placement = "H"   # <<< molto importante
)

tab_q
```

