
---
title: "Model Specification and Diagnostic Analysis: Stationarity, Persistence, and Endogeneity in the Baseline TWFE Framework"
author: "Sofia Perondi"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: false
---

```{r setup, include=FALSE}
# ========== SETUP ==========
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(plm)
library(lmtest)
library(sandwich)
library(stargazer)
```

The baseline specification aimed to explain the growth of real public expenditure in EU countries between 2011 and 2021 through a Two-Way Fixed Effects (TWFE) framework.
Although methodologically correct, the initial model presented several weaknesses: it explained only 28.7% of within-country variation, and only two variables — old-age dependency ratio and public debt — were statistically significant.
Upon diagnostic review, a critical omitted variable problem was identified.
The model lacked a key macroeconomic control: economic growth.
Without controlling for the business cycle, the baseline equation suffered from omitted variable bias, as public spending is strongly influenced by GDP dynamics through both revenue and fiscal policy channels.
To correct this, the recession dummy was replaced with a continuous measure of real GDP growth (gdp_growth_real).
This adjustment improved both theoretical coherence and empirical performance, as the new specification aligns the explanatory variable with the dependent variable (both expressed in real terms).

```{r}
# === 1️⃣ Import data ===

# Path assoluto senza spazi problematici
DATA_PATH <- file.path("~/Desktop/Tesina_IV/Data_analysis/Downloaded_data/FINAL_ANALYSIS_DATASET_2006_2021.csv")

# Oppure: usa il path relativo se il file è nella stessa cartella del tuo .Rmd
# DATA_PATH <- "FINAL_ANALYSIS_DATASET_2006_2021.csv"

stopifnot(file.exists(path.expand(DATA_PATH)))  # Controlla che esista prima di procedere
final_dataset <- read_csv(DATA_PATH, show_col_types = FALSE) %>%
  filter(tender_year >= 2011)

# === 2️⃣ Build panel ===

macro_updated <- final_dataset %>%
group_by(buyer_country, tender_year) %>%
summarise(
total_expenditure_growth_real = first(total_expenditure_growth_real),
gdp_growth_real = first(gdp_growth_real),
gdp_growth = first(gdp_growth),
unemp_rate = first(unemp_rate),
old_dep = first(old_dep),
pop_density = first(pop_density),
gov_debt_pct_gdp = first(gov_debt_pct_gdp),
corruption_control = first(corruption_control),
gov_state_market = first(gov_state_market),
.groups = "drop"
)

mp_updated <- pdata.frame(macro_updated, index = c("buyer_country", "tender_year"))

# === 3️⃣ Estimate models ===

fml_baseline <- total_expenditure_growth_real ~ unemp_rate + old_dep + pop_density +
gov_debt_pct_gdp + corruption_control + gov_state_market + factor(tender_year)
fml_improved <- update(fml_baseline, . ~ gdp_growth_real + .)
fml_nominal <- update(fml_baseline, . ~ gdp_growth + .)

m_twfe_baseline <- plm(fml_baseline, data = mp_updated, model = "within")
m_twfe_improved <- plm(fml_improved, data = mp_updated, model = "within")
m_twfe_nominal <- plm(fml_nominal, data = mp_updated, model = "within")

# === 4️⃣ Clustered SE ===

se_baseline <- sqrt(diag(vcovHC(m_twfe_baseline, type = "HC1", cluster = "group")))
se_improved <- sqrt(diag(vcovHC(m_twfe_improved, type = "HC1", cluster = "group")))
se_nominal <- sqrt(diag(vcovHC(m_twfe_nominal, type = "HC1", cluster = "group")))
```


```{r}
stargazer(m_twfe_nominal, m_twfe_improved,
type = "latex",
title = "Comparison of TWFE Models: Nominal vs Real GDP Growth",
dep.var.labels = "Real Public Expenditure Growth",
column.labels = c("Nominal GDP", "Real GDP"),
covariate.labels = c("GDP Growth", "Unemployment Rate", "Old-age Dependency",
"Population Density", "Public Debt (% GDP)",
"Corruption Control", "Government Orientation (Market)",
"Year Fixed Effects"),
add.lines = list(
c("R² (within)",
sprintf("%.3f", summary(m_twfe_nominal)$r.squared["rsq.within"]),
sprintf("%.3f", summary(m_twfe_improved)$r.squared["rsq.within"]))
),
omit.stat = c("f", "ser"),
notes = "Cluster-robust standard errors by country in parentheses.",
notes.align = "l")
```


Including real GDP growth materially improves the model:
- the within–R² increases (see Table 1),
- more coefficients become statistically meaningful,
- and the signs align with macro-fiscal theory.
Real GDP growth emerges as the main determinant of expenditure growth, while market-oriented governments are associated with more restrained spending dynamics.

To verify robustness, a comparison was performed between specifications using nominal and real GDP growth.
Although both series are highly correlated (ρ ≈ 0.87), the real measure provides a more reliable macroeconomic signal, being unaffected by inflationary effects.
As shown in Table 1, the real GDP specification yields slightly higher explanatory power and more stable coefficients.


