---
title: "Public Procurement Sectoral Analysis"
author: "Sofia"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
library(knitr)

# Load your data
gppd_complete <- read.csv("gppd_outsourcing_data_FINAL.csv")

# Prepare analysis dataset
gppd_analysis <- gppd_complete %>%
  filter(!is.na(tender_digiwhist_price) & tender_digiwhist_price > 0) %>%
  mutate(
    tender_supplytype = toupper(trimws(tender_supplytype)),
    sector = case_when(
      tender_supplytype == "SERVICES" ~ "Services",
      tender_supplytype == "SUPPLIES" ~ "Supplies", 
      tender_supplytype == "WORKS" ~ "Works",
      TRUE ~ "Other"
    ),
    log_value = log(tender_digiwhist_price + 1),
    sector = as.factor(sector),
    buyer_country = as.factor(buyer_country),
    tender_year = as.factor(tender_year)
  )

# Sector distribution
sector_dist <- gppd_analysis %>%
  group_by(sector) %>%
  summarise(
    n_contracts = n(),
    total_value = sum(tender_digiwhist_price, na.rm = TRUE) / 1e9,
    avg_value = mean(tender_digiwhist_price, na.rm = TRUE) / 1e6,
    .groups = 'drop'
  ) %>%
  mutate(share = n_contracts / sum(n_contracts) * 100)

# Mixed model
library(lme4)
model1 <- lmer(log_value ~ 1 + (1|sector) + (1|buyer_country) + (1|tender_year), 
               data = gppd_analysis)
vc <- as.data.frame(VarCorr(model1))
total_variance <- sum(vc$vcov)
vc$variance_pct <- vc$vcov / total_variance * 100

# Sector concentration
sector_concentration <- gppd_analysis %>%
  group_by(buyer_country, sector) %>%
  summarise(sector_value = sum(tender_digiwhist_price, na.rm = TRUE), .groups = 'drop') %>%
  group_by(buyer_country) %>%
  mutate(
    total_country = sum(sector_value),
    sector_share = sector_value / total_country
  ) %>%
  summarise(
    hhi = sum(sector_share^2) * 10000,
    n_sectors = n_distinct(sector),
    dominant_sector = sector[which.max(sector_share)],
    dominant_share = max(sector_share),
    .groups = 'drop'
  )

# Sector trends
sector_trends <- gppd_analysis %>%
  group_by(tender_year, sector) %>%
  summarise(
    total_value = sum(tender_digiwhist_price, na.rm = TRUE),
    n_contracts = n(),
    .groups = 'drop'
  ) %>%
  group_by(tender_year) %>%
  mutate(
    annual_total = sum(total_value),
    sector_share = total_value / annual_total
  )

# Plots
sector_plot <- sector_dist %>%
  mutate(
    total_value = total_value * 1e9,
    avg_value = avg_value * 1e6
  )

p1 <- ggplot(sector_plot, aes(x = reorder(sector, -share), y = n_contracts)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::comma(n_contracts)), vjust = -0.3, size = 3) +
  labs(title = "Number of Contracts by Sector (EU-27)",
       x = "Sector", y = "Contracts") +
  theme_minimal()

p2 <- ggplot(sector_plot, aes(x = reorder(sector, -total_value), y = total_value/1e12)) +
  geom_col(fill = "darkorange") +
  geom_text(aes(label = paste0("$", round(total_value/1e12, 1), "T")),
            vjust = -0.3, size = 3) +
  labs(title = "Total Value of Contracts by Sector (EU-27)",
       x = "Sector", y = "Contract Value (USD Trillions)") +
  theme_minimal()

contracts_trends <- gppd_analysis %>%
  group_by(tender_year, sector) %>%
  summarise(n_contracts = n(), .groups = 'drop')

p3 <- ggplot(contracts_trends, aes(x = as.numeric(as.character(tender_year)), 
                                   y = n_contracts, color = sector)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  labs(title = "Annual Distribution of Contracts by Sector (EU-27)",
       x = "Year", y = "Number of Contracts") +
  theme_minimal()

top10_countries <- gppd_analysis %>%
  group_by(buyer_country) %>%
  summarise(total_value = sum(tender_digiwhist_price, na.rm = TRUE)) %>%
  top_n(10, total_value) %>%
  pull(buyer_country)

p4 <- gppd_analysis %>%
  filter(buyer_country %in% top10_countries) %>%
  ggplot(aes(x = log_value, fill = buyer_country)) +
  geom_density(alpha = 0.4) +
  labs(title = "Distribution of Contract Values (Top-10 EU Countries)",
       x = "Log Contract Value", y = "Density") 
  theme_minimal()

```

```{r sector-table}
sector_dist %>%
  mutate(
    total_value = round(total_value, 2),
    avg_value = round(avg_value, 2),
    share = round(share, 1)
  ) %>%
  kable(
    caption = "Sector Distribution in EU Public Procurement",
    col.names = c("Sector", "Contracts", "Total Value (€B)", "Avg Value (€M)", "Share (%)")
  )

```

```{r variance-table}
vc %>%
  mutate(
    vcov = round(vcov, 3),
    variance_pct = round(variance_pct, 1)
  ) %>%
  kable(
    caption = "Variance Components in Log Contract Values",
    col.names = c("Group", "Effect 1", "Effect 2", "Variance", "SD", "Variance (%)")
  )

```

```{r concentration-table}

sector_concentration %>%
  mutate(
    hhi = round(hhi, 0),
    dominant_share = round(dominant_share * 100, 1)
  ) %>%
  kable(
    caption = "Sectoral Concentration by Country",
    col.names = c("Country", "HHI", "# Sectors", "Dominant Sector", "Dominant Share (%)")
  )

```

```{r trends-table}
sector_trends %>%
  filter(tender_year %in% c("2007", "2010", "2015", "2020")) %>%
  mutate(
    total_value = round(total_value / 1e9, 2),
    sector_share = round(sector_share * 100, 1)
  ) %>%
  kable(
    caption = "Sector Trends Over Time (Selected Years)",
    col.names = c("Year", "Sector", "Value (€B)", "Contracts", "Annual Total (€)", "Share (%)")
  )

```

```{r plot-p1}
p1

```

```{r plot-p2}
p2

```

```{r plot-p3}
p3

```

```{r plot-p4}
p4

```

This report analyzes over 2 million public procurement contracts across 27 EU countries from 2002 to 2021. Using `tender_supplytype`, contracts were classified into Services, Supplies, and Works.

- **Sector Distribution**: Supplies dominate both in volume and value, accounting for over half of all contracts. Services and Works follow, with Works showing higher average contract values.
- **Variance Decomposition**: Mixed-effects modeling reveals that country-level differences explain 23% of the variation in contract value, sector explains 7%, and year explains 4.5%. The remaining 65% is residual, reflecting buyer-specific and unobserved factors.
- **Sectoral Concentration**: Some countries (e.g., Bulgaria) are highly specialized in one sector, while others maintain a more balanced procurement portfolio. The average Herfindahl-Hirschman Index (HHI) is 5090, indicating moderate concentration.
- **Trends Over Time**: Works contracts peaked in 2007, likely due to infrastructure investment cycles. Supplies have grown steadily, reflecting ongoing material needs across sectors.
- **Top Countries**: The distribution of contract values varies widely across top buyers, with some countries showing tighter clustering and others more dispersion.

This analysis provides a foundation for understanding procurement behavior across the EU and identifying patterns of specialization, investment, and strategic variation.
  
  
