---
title: "Full Model – Demeaned Public Expenditure and Outsourcing"
author: "Sofia Perondi"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: false
fontsize: 11pt
---

# Data, Variable Construction, and Specification
This report replicates the analysis of the main model but uses a demeaned dependent variable for public expenditure (log primary expenditure demeaned within each country).
This allows a stricter comparison between within-country deviations of public expenditure and outsourcing intensity, removing structural country-level differences.
The structure, models, and tests are identical to the baseline report to enable direct comparison.

```{r setup, include=FALSE}
library(tidyverse)
library(fixest)
library(modelsummary)
library(plm)
library(car)
library(knitr)
library(kableExtra)
options(knitr.table.format = "latex")

setwd("~/Desktop/Tesina IV /Data analysis /Downloaded data")

# Load main macro dataset and auxiliary data

df <- read_csv("macro_primary_expenditure_with_growth_controls.csv", show_col_types = FALSE)

expprim <- read_csv("eurostat_primary_expenditure_real.csv", show_col_types = FALSE) %>%
select(geo, year, primary_real, gdp_deflator)

final <- read_csv("FINAL_ANALYSIS_DATASET_2006_2021.csv", show_col_types = FALSE) %>%
select(buyer_country, tender_year, gdp_nominal) %>%
distinct()

df <- df %>%
left_join(expprim, by = c("buyer_country" = "geo", "tender_year" = "year")) %>%
left_join(final, by = c("buyer_country", "tender_year")) %>%
mutate(
gdp_real = if_else(!is.na(gdp_nominal) & !is.na(gdp_deflator) & gdp_deflator > 0,
gdp_nominal / (gdp_deflator / 100), NA_real_)
)
```

## Transformations and Demeaned Variables
We apply log transformations to continuous variables and demean them by country where relevant.
The dependent variable is now log real primary expenditure demeaned within each country (ln_spesa_dm), ensuring that estimation captures within-country dynamics only.


```{r, include=FALSE}
df <- df %>%
filter(!is.na(primary_real), primary_real > 0,
!is.na(avg_contract_price), avg_contract_price > 0,
!is.na(unemp_rate), unemp_rate > 0,
!is.na(pop_density), pop_density > 0,
!is.na(gov_debt_pct_gdp), gov_debt_pct_gdp > 0,
!is.na(gdp_real), gdp_real > 0) %>%
mutate(
ln_spesa = log(primary_real),
ln_valmean = log(avg_contract_price)
) %>%
group_by(buyer_country) %>%
mutate(
ln_spesa_dm = ln_spesa - mean(ln_spesa, na.rm = TRUE),
X = ln_valmean - mean(ln_valmean, na.rm = TRUE)
) %>%
ungroup() %>%
mutate(
ln_unemp_rate = log(unemp_rate),
ln_pop_density = log(pop_density),
ln_gov_debt = log(gov_debt_pct_gdp),
ln_gdp_real = log(gdp_real)
)
```

```{r,echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
vars <- tribble(
~Variable, ~Description, ~Transformation,
"ln_spesa_dm", "Log real primary expenditure demeaned within country", "log(level) demeaned",
"X", "Outsourcing intensity: log mean contract value demeaned within country", "log-level deviation",
"ln_unemp_rate", "Unemployment rate (% labour force)", "log(level)",
"ln_pop_density", "Population per km²", "log(level)",
"ln_gov_debt", "Government debt as % of GDP", "log(level)",
"corruption_control", "WGI Control of Corruption (−2.5 to +2.5)", "raw index",
"gov_state_market", "Government state–market orientation (Manifesto index)", "raw index",
"ln_gdp_real", "Real GDP (nominal GDP / GDP deflator)", "log(level)",
"recession", "Recession dummy (1 = recession year)", "0/1",
"unemp_growth", "Unemployment rate growth", "% change",
"pop_growth", "Population density growth", "% change",
"gdp_growth", "Real GDP growth", "% change"
)

n <- nrow(vars); cut <- ceiling(n/2)
vars_part1 <- vars %>% slice(1:cut)
vars_part2 <- vars %>% slice((cut+1):n)

kable(vars_part1, booktabs = TRUE, caption = "Variable descriptions (part 1)") %>%
kable_styling(font_size = 7, latex_options = c("scale_down", "HOLD_position"))

kable(vars_part2, booktabs = TRUE, caption = "Variable descriptions (part 2)") %>%
kable_styling(font_size = 7, latex_options = c("scale_down", "HOLD_position"))
```

# Baseline and Comparative Models
We estimate four specifications to progressively account for unobserved heterogeneity: pooled OLS, country fixed effects, year fixed effects, and two-way fixed effects (TWFE). Cluster-robust errors by country address potential serial correlation within panels. The TWFE model serves as our baseline specification.

```{r,echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
m_pool <- feols(
ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density + ln_gov_debt +
corruption_control + gov_state_market + ln_gdp_real,
data = df, cluster = ~ buyer_country
)

m_country <- feols(
ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density + ln_gov_debt +
corruption_control + gov_state_market + ln_gdp_real | buyer_country,
data = df, cluster = ~ buyer_country
)

m_year <- feols(
ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density + ln_gov_debt +
corruption_control + gov_state_market + ln_gdp_real | tender_year,
data = df, cluster = ~ buyer_country
)

m_twfe <- feols(
ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density + ln_gov_debt +
corruption_control + gov_state_market + ln_gdp_real |
buyer_country + tender_year,
data = df, cluster = ~ buyer_country
)

etable(
list(
"Pooled OLS (no FE)" = m_pool,
"Country FE" = m_country,
"Year FE" = m_year,
"Two-way FE (TWFE)" = m_twfe
),
se = "cluster",
cluster = "buyer_country",
fitstat = ~ n + r2,
tex = TRUE,
digits = 4,
headers = c( "Pooled OLS (no FE)", 
              "Country FE", 
              "Year FE", 
              "Two-way FE (TWFE)")
) %>%
  cat() 
```

# Multicollinearity
To ensure that explanatory variables do not overlap excessively, we compute Variance Inflation Factors (VIF). All values are comfortably below standard concern thresholds (typically VIF < 5), indicating that multicollinearity is not a significant issue in the baseline specification.

```{r,echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
m_noFE <- lm(
ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density +
ln_gov_debt + corruption_control + gov_state_market + ln_gdp_real,
data = df
)

vif_tbl <- tibble(Variable = names(vif(m_noFE)), VIF = as.numeric(vif(m_noFE)))

n_v <- nrow(vif_tbl); cut_v <- ceiling(n_v/2)
vif_part1 <- vif_tbl %>% slice(1:cut_v)
vif_part2 <- vif_tbl %>% slice((cut_v+1):n_v)

kable(vif_part1, booktabs = TRUE, caption = "Multicollinearity (part 1)") %>%
kable_styling(latex_options = c("scale_down", "HOLD_position"))

kable(vif_part2, booktabs = TRUE, caption = "Multicollinearity (part 2)") %>%
kable_styling(latex_options = c("scale_down", "HOLD_position"))
```

# Variance decomposition 
We decompose each variable’s total variance into between-country and within-country components. 

```{r,echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
pdata <- pdata.frame(df, index = c("buyer_country", "tender_year"))
vars_to_check <- c("ln_spesa_dm","X","ln_unemp_rate","ln_pop_density",
"ln_gov_debt","corruption_control","gov_state_market","ln_gdp_real")

get_var_components <- function(v) {
x <- pdata[[v]]
means_country <- tapply(x, pdata$buyer_country, mean, na.rm = TRUE)
between_var <- var(means_country, na.rm = TRUE)
within_var  <- mean(tapply(x, pdata$buyer_country, var, na.rm = TRUE), na.rm = TRUE)
total_var   <- between_var + within_var
tibble(variable = v,
total_var = total_var,
between_var = between_var,
within_var = within_var,
between_share = between_var/total_var,
within_share  = within_var/total_var)
}

var_summary <- bind_rows(lapply(vars_to_check, get_var_components)) %>%
arrange(within_share)

kable(var_summary %>% mutate(across(where(is.numeric), ~round(., 6))),
booktabs = TRUE, caption = "Variance decomposition: within vs between share.") %>%
kable_styling(full_width = FALSE, latex_options = c("scale_down", "HOLD_position"))
```

# Robustness Checks
We re-estimate the TWFE model under two alternative specifications: one replacing real GDP with a recession dummy, and another replacing log-level controls with growth-rate measures. 

The direction and magnitude of the outsourcing coefficient remain stable, supporting the robustness of our main findings.

```{r robustness_table, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
# --- Safe path for GDP growth file ---
ctrl_file <- "~/Desktop/Tesina IV /Data analysis /Downloaded data/all_controls_complete_2006_2021.csv"

if (file.exists(ctrl_file)) {
  ctrl_gdp <- read_csv(ctrl_file, show_col_types = FALSE) %>%
    transmute(
      buyer_country = geo,
      tender_year   = year,
      gdp_growth    = gdp_growth
    ) %>%
    filter(tender_year >= 2011, tender_year <= 2021)
  
  df <- df %>%
    left_join(ctrl_gdp, by = c("buyer_country", "tender_year"))
} else {
  warning("⚠️ GDP growth file not found. Skipping merge.")
}

# --- Define robustness models ---
m_twfe_rec <- feols(
  ln_spesa_dm ~ X + ln_unemp_rate + ln_pop_density + ln_gov_debt +
    corruption_control + gov_state_market + recession |
    buyer_country + tender_year,
  data = df, cluster = ~ buyer_country
)

m_twfe_growth <- feols(
  ln_spesa_dm ~ X + unemp_growth + pop_growth + gdp_growth +
    gov_debt_pct_gdp + corruption_control + gov_state_market |
    buyer_country + tender_year,
  data = df, cluster = ~ buyer_country
)

# --- Robustness etable in LaTeX ---
etable(
  list(
    "Baseline (TWFE log GDP)"   = m_twfe,
    "TWFE + Recession Dummy"    = m_twfe_rec,
    "TWFE with Growth Controls" = m_twfe_growth
  ),
  se = "cluster",
  cluster = "buyer_country",
  fitstat = ~ n + r2,
  digits = 4,
  tex = TRUE,
  title = "Robustness Check: Two-way Fixed Effects models"
)
```